---
title: "Class 13 - RNAseq mini project"
author: "Hetian Su"
format: pdf
editor: visual
---

```{r, message=FALSE}
# load the DESeq2 package
library(DESeq2)
```

```{r}
# load count and metadata for DEseq analysis
metaFile <- 'GSE37704_metadata.csv'
countFile <- 'GSE37704_featurecounts.csv'

colData <- read.csv(metaFile, row.names = 1)
head(colData)

countData <- read.csv(countFile, row.names = 1)
head(countData)
```

### Q1

```{r}
# remove the first length column from count data
countData <- as.matrix(countData[,-1])
head(countData)
```

### Q2

```{r}
# remove gene entries that have 0 in each sample
countData <- countData[rowSums(countData)!=0, ]
head(countData)
```

PCA quality control on the dataset to see if different conditions are separable.

```{r}
pca <- prcomp(t(countData), scale. = T)
```

```{r}
plot(pca$x[,1], pca$x[,2], col=as.factor(colData$condition))
```

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)

dds
```

### Q3

```{r}
res <- results(dds)
summary(res)
```

## Visualization

### Q4

```{r}
mycols <- rep('gray', nrow(res))
mycols[abs(res$log2FoldChange) > 2] <- 'red'
inds <- (res$padj<0.01)&(abs(res$log2FoldChange)>2)
mycols[inds] <- 'blue'
plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

## Add Annotation

### Q5

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column='SYMBOL',
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype='ENSEMBL',
                    column='GENENAME',
                    multiVals="first")

head(res, 10)
```

### Q6

```{r}
# reorder the annotated gene data by adjusted p-values
res <- res[order(res$padj),]
write.csv(res, 'deseq_results.csv')
```

## Pathway Analysis

```{r}
# BiocManager::install( c("pathview", "gage", "gageData") )
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

kegg.sets.hs <- kegg.sets.hs[sigmet.idx.hs]
head(kegg.sets.hs,3)
```

```{r}
# extract a named vector of fold changes to use as input to pathway mapping
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# map the extracted fold changes to pathways defined in the kegg.sets.hs, a portion of pathway database contianing signaling and metabolism pathways
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
head(keggres$less)
```

```{r}
# we find that the cell cycle pathway is the most down-regulated, map gene fold changes to this pathway
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

Try to process the 5 most upregulated pathways

```{r}
keggrespathways <- row.names(keggres$greater)[1:5]
keggresids <- substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r, message=FALSE}
pathview(gene.data = foldchanges, pathway.id = keggresids, species = 'hsa')
```

### Q7 Repeat for the 5 most down regulated pathways

```{r}
keggresdownpathways <- row.names(keggres$less)[1:5]
keggresdownids <- substr(keggresdownpathways, start=1, stop=8)
pathview(gene.data = foldchanges, pathway.id = keggresdownids, species = 'hsa')
```

## GO Term Analysis

```{r}
data(go.sets.hs) #all GO terms
data(go.subs.hs) #GO terms realted to BP, CC, MF

gobpsets <- go.sets.hs[go.subs.hs$BP]
gobpres <- gage(foldchanges, gsets = gobpsets, same.dir = T)
```

```{r}
lapply(gobpres, head)
```

## Reactome Analysis

```{r}
# see how many significantly changed genes are there in total
siggene <- res[res$padj<=0.05 & !is.na(res$padj),'symbol']
print(paste('Total number of significantly changed genes is:', length(siggene)))
```

```{r}
write.table(siggene, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

### Q8

The Endosomal/Vacuolar pathway has the most significant entities p-value. The results do not match well the KEGG results. It seems that Reactome Database has overall a more detailed definition of "reaction", such that the same general type reaction in KEGG could have more versions, or subcategories, in Reactome. Here, KEGG analysis was restricted to signaling and metabolism pathways, whereas we seems to have not restricted the search in Reactome.
