---
title: "Pertussis Mini Project"
author: "Hetian Su"
format: pdf
editor: visual
---

## Q1

```{r}
# install.packages('datapasta')
library(datapasta)
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,2019L),
  No..Reported.Pertussis.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617)
)
```

```{r}
library(ggplot2)
plot <- ggplot(cdc)+
  aes(Year, No..Reported.Pertussis.Cases)+
  geom_point()+
  geom_line()+
  labs(title='Pertussis Cases by Year (1922-2019)', x='Year', y='Number of Cases')
plot
```

## Q2

```{r}
plot <- plot + geom_vline(xintercept = 1946, color='blue', linetype='dashed') + geom_vline(xintercept = 1996, color='red', linetype='dashed')
plot
```

## Q3

**It can be seen that after the introduction of aP vaccine, the number of cases started to increase again. It is possible that the bacterial evolution could more easily escape the protection provided by the aP vaccine.**

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject)
```

## Q4

```{r}
sum(subject$infancy_vac=='wP')
sum(subject$infancy_vac=='aP')
```

**There are 47 aP vaccinated subjects, and 49 wP vaccinated subjects.**

## Q5

```{r}
sum(subject$biological_sex=='Male')
sum(subject$biological_sex=='Female')
```

**There are 30 males and 66 females.**

## Q6

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
# install.packages('lubridate')
library(lubridate)
```

## Q7

```{r}
wP_age <- time_length(today()-ymd(subject$year_of_birth[subject$infancy_vac=='wP']), 'years')
aP_age <- time_length(today()-ymd(subject$year_of_birth[subject$infancy_vac=='aP']), 'years')
```

```{r}
mean(wP_age)
mean(aP_age)
t.test(wP_age, aP_age)
```

**The mean age of wP vaccinated subjects is 36 years old, and that of aP vaccinated subjects is 25 years old. They are significantly different under 2-sample t test.**

## Q8

```{r}
age_at_boost <- time_length(ymd(subject$date_of_boost)-ymd(subject$year_of_birth), 'years')
head(age_at_boost)
```

## Q9.a

```{r}
subject$age <- today()-ymd(subject$year_of_birth)
ggplot(subject)+
  aes(time_length(age, 'years'), fill=as.factor(infancy_vac))+
  geom_histogram(show.legend = FALSE)+
  facet_wrap(vars(infancy_vac), nrow = 2)
```

**The distribution shown in the plot here shows that the 2 groups have significantly different ages.**

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = TRUE)
```

## Q9.b

```{r}
library(dplyr)
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

## Q10

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

## Q11

```{r}
table(abdata$isotype)
```

## Q12

```{r}
table(abdata$visit)
```

**There are a lot fewer specimens collected on the 8th visit.**

```{r}
ig1 <- abdata%>%filter(isotype=='IgG1', visit!=8)
head(ig1)
```

## Q13

```{r}
ggplot(ig1)+
  aes(MFI,antigen)+
  geom_boxplot()+
  facet_wrap(vars(visit), nrow = 2)
```

## Q14

**Judging by the change of mean and distribution shown by boxplots over visits, TT, PRN, FIM2/3, FHA levels increased overtime, with FIM2/3 being the most significant.**

## Q15

```{r}
filter(ig1, antigen=="Measles") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(ig1, antigen=='FIM2/3')%>%
  ggplot()+
  aes(MFI, col=infancy_vac)+
  geom_boxplot(show.legend = FALSE)+
  facet_wrap(vars(visit))+
  theme_bw()
```

## Q16

**As compared to the negative control, both wP and aP vaccine induced high level production of FIM2/3. Also notably, overtime the level of FIM2/3 induced by aP maintains but that by wP decreases.**

## Q17

**Overtime, antigen produced in response to aP increases and stays at high levels, thus eventually exceeding the level of that induced by wP which decreases after some time.**

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
ssrna <- inner_join(rna, meta)
```

## Q18

```{r}
ggplot(ssrna)+
  aes(visit, tpm, group=subject_id)+
  geom_point()+
  geom_line(alpha=0.2)
```

## Q19

**The expression of this gene tends to increase abruptly and peak at the 4th visit.**

## Q20

**The peaking behavior matches closely with FIM2/3 pattern induced by wP, whereas FIM2/3 induced by aP seem to be able to stay at rather high levels.**

```{r}
ggplot(ssrna)+
  aes(tpm, col=infancy_vac)+
  geom_boxplot()+
  facet_wrap(vars(visit))
```

```{r}
ssrna%>%filter(visit==4)%>%ggplot()+
  aes(tpm, col=infancy_vac)+
  geom_density()+
  geom_rug()
```
